---
title: "Project by Stephanie Miller & Duong Le"
author: "Duong Le"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)
```

## Loading packages 
```{r}
library(tidyverse)
library(gridExtra)
library(janitor)
library(janitor)
library(colorspace)
library(scales)
library(knitr)
library(kableExtra)
library(formattable)
library(lubridate)
library(dplyr)
```

## Loading and transforming data 
```{r}
loan_data <- read.csv("loan_3.csv", header = TRUE)
```

```{r}
head(loan_data,20)
```

### Renaming columns
```{r}
names(loan_data)[names(loan_data) == "v1"] <- "Credit_Score"
names(loan_data)[names(loan_data) == "v2"] <- "First_Payment_Date"
names(loan_data)[names(loan_data) == "v3"] <- "First_Time_Homebuyer_Flag"
names(loan_data)[names(loan_data) == "v4"] <- "Maturity_Date"
names(loan_data)[names(loan_data) == "v5"] <- "MSA"
names(loan_data)[names(loan_data) == "v6"] <- "MI%"
names(loan_data)[names(loan_data) == "v7"] <- "Number_of_Units"
names(loan_data)[names(loan_data) == "v8"] <- "Occupancy_Status"
names(loan_data)[names(loan_data) == "v9"] <- "CLTV"
names(loan_data)[names(loan_data) == "v10"] <- "DTI_Ratio"
names(loan_data)[names(loan_data) == "v11"] <- "Original_UPB"
names(loan_data)[names(loan_data) == "v12"] <- "Original_LTV"
names(loan_data)[names(loan_data) == "v13"] <- "Original_Interest"
names(loan_data)[names(loan_data) == "v14"] <- "Channel"
names(loan_data)[names(loan_data) == "v15"] <- "PPM"
names(loan_data)[names(loan_data) == "v16"] <- "Product_Type"
names(loan_data)[names(loan_data) == "v17"] <- "State"
names(loan_data)[names(loan_data) == "v18"] <- "Property_Type"
names(loan_data)[names(loan_data) == "v19"] <- "Postal_Code"
names(loan_data)[names(loan_data) == "v20"] <- "Loan_Sequence_Number"
names(loan_data)[names(loan_data) == "v21"] <- "Loan_Purpose"
names(loan_data)[names(loan_data) == "v22"] <- "Original_Loan_Term"
names(loan_data)[names(loan_data) == "v23"] <- "Number_of_Borrowers"
names(loan_data)[names(loan_data) == "v24"] <- "Seller_Name"
names(loan_data)[names(loan_data) == "v25"] <- "Service_Name"
names(loan_data)[names(loan_data) == "v26"] <- "Super_Conforming_Flag"
names(loan_data)[names(loan_data) == "v27"] <- "Risk_Level"

```

```{r}
str(loan_data)
```

```{r}
summary(loan_data)
```

### Formatting variables
```{r}
#First Time Homebuyer (factor)
loan_data$First_Time_Homebuyer_Flag <- as.factor(loan_data$First_Time_Homebuyer_Flag)

#PPM (factor)
loan_data$PPM <- as.factor(loan_data$PPM)

#Postal Code (factor)
loan_data$Postal_Code <- as.factor(loan_data$Postal_Code)

#Loan Sequence Number (numeric)
loan_data$Loan_Sequence_Number <- as.numeric(loan_data$Loan_Sequence_Number)

#Number of Borrowers (factor)
loan_data$Number_of_Borrowers <- as.factor(loan_data$Number_of_Borrowers)

#Risk level (factor)
loan_data$Risk_Level <- as.factor(loan_data$Risk_Level)

```



## Base EDA Step 1: Uni-variate non-graphical
+ Unordered but relatively clean data, each column is of variable type, each row is an observation of a loan request.
+ Potential redundancy in column 9 ("Original combined LTV") and column 12 ("Original LTV"). 
+ Concerns include the **amount of information that is not applicable or missing** (see column 3 "Homebuyer Flag"), which might influence what variables would be include in predictive modelling.
+ There are also high number of levels in the categorical variables (see postal code and property state), which might be difficult to visualize.
+ Potential areas of interest: **debt-to-income ratio** (preferrably in the lower %, usually in the 20s to 40s), **LTV** (preferrably in the lower %, usually in the 80s).  

## Base EDA Step 2: Uni-variate graphical 

### Categorical
Note: Product type is not graphed as there's only one type (FRM)
```{r}
grid.arrange(
  
#First Time Homebuyer Flag
ggplot(data = loan_data, mapping = aes(x = First_Time_Homebuyer_Flag)) + geom_bar(),

#Occupancy Status
ggplot(data = loan_data, mapping = aes(x = Occupancy_Status)) + geom_bar(),

#Channel
ggplot(data = loan_data, mapping = aes(x = Channel)) + geom_bar(), 

#PPM 
ggplot(data = loan_data, mapping = aes(x = PPM)) + geom_bar(), 

#Property Type
ggplot(data = loan_data, mapping = aes(x = Property_Type)) + geom_bar(), 

#Loan Purpose
ggplot(data = loan_data, mapping = aes(x = Loan_Purpose)) + geom_bar(), 

#Number of Borrowers
ggplot(data = loan_data, mapping = aes(x = Number_of_Borrowers)) + geom_bar(), 

#Conforming Flag
ggplot(data = loan_data, mapping = aes(x = Super_Conforming_Flag)) + geom_bar(),

ncol = 2)

```

```{r}
#Property State 
ggplot(data = loan_data, mapping = aes(x = State)) + geom_bar()

#Postal Code
ggplot(data = loan_data, mapping = aes(x = Postal_Code)) + geom_bar()
```
Note: Variables that indicate location e.g. postal code, state, or MSA is not necessarily conducive to determing loan eligibility. If there's interest in which areas are requesting the most loans, the top 10 areas could be graphed for visualization. 

### Quantitative
Note: Graphing variables of interest only. 

+ Adjusting Outliers & Missing Values

```{r}
#Credit Score
loan_data$Credit_Score[loan_data$Credit_Score == 9999] <- NA

ggplot(data = loan_data, mapping = aes(x = Credit_Score)) +
    geom_histogram()
```

```{r}
#CLTV
loan_data$CLTV[loan_data$CLTV == 999] <- NA

ggplot(data = loan_data, mapping = aes(x = CLTV)) +
    geom_histogram()
```

```{r}
#DTI
loan_data$DTI_Ratio[loan_data$DTI_Ratio == 999] <- NA

ggplot(data = loan_data, mapping = aes(x = DTI_Ratio)) +
    geom_histogram()
```

```{r}
#Original Interest Rate
ggplot(data = loan_data, mapping = aes(x = Original_Interest)) +
    geom_histogram()
```

```{r}
# Original Loan Term
ggplot(data = loan_data, mapping = aes(x = Original_Loan_Term)) +
    geom_histogram()
```


## Base EDA Step 3: Multi-variate non-graphical 
### Categorical

+ First Time Homebuyer Flag.
+ Occupancy status.
+ PPM.
+ Property type.
+ Loan purpose.
+ Number of borrowers.
+ Conforming flag.

```{r}
#Flag & Occupancy Status
loan_data %>%
  tabyl(First_Time_Homebuyer_Flag, Occupancy_Status) %>%
  adorn_totals(where = c("row", "col")) %>% 
  adorn_percentages(denominator = "all") %>%
  adorn_rounding(2)
```

+ The majority of customers are borrowing to pay off their primary residence.
+ There are more non-first-time home buyers borrowing than first-timers. 

```{r}
#Flag & Number of Borrowers
loan_data %>%
  tabyl(First_Time_Homebuyer_Flag, Number_of_Borrowers) %>%
  adorn_totals(where = c("row", "col")) %>% 
  adorn_percentages(denominator = "all") %>%
  adorn_rounding(2)
```

```{r}
#Flag and Conformity
loan_data %>%
  tabyl(First_Time_Homebuyer_Flag, Super_Conforming_Flag) %>%
  adorn_totals(where = c("row", "col")) %>% 
  adorn_percentages(denominator = "all") %>%
  adorn_rounding(2)
```

+ Non-conforming loans are rarely offered; instances of offer is for primary residence rather than investment property or second home.
+ Only 5 percent of first-time homeowners get offered non-conforming loans.
+ First time homeowners borrow for primary residence rather as investment properties

### Quantitative 

```{r}
loan_data %>%
  select_if(is.numeric) %>%
  cor() %>%
  round(2) %>%
  kable()
```

+ 0.63 correlation original interest and original loan term
+ 0.29 correlation original loan term and original LTV
+ 0.16 correlation original UPB and MSA
+ 0.19 original UPB and original loan term


## Base EDA Step 4: Multi-variate graphical and Detailed Analysis
### Categorical 
+ First Time Homebuyer Flag.
+ Occupancy status.
+ Property type.
+ Loan purpose.
+ Number of borrowers.
+ Conforming flag.
+ Risk level.


```{r}
# First Time Homebuyer & Property Type
revised_hf <- c("Y", "N")
loan_data %>%
  filter(First_Time_Homebuyer_Flag %in% revised_hf) %>%
  ggplot(mapping = aes(x = First_Time_Homebuyer_Flag, fill = Property_Type)) +
  geom_bar(position = "dodge")
```


```{r}
# First Time Homebuyer & Number of Borrowers
loan_data %>%
    filter(First_Time_Homebuyer_Flag %in% revised_hf) %>%
  ggplot(mapping = aes(x = First_Time_Homebuyer_Flag, fill = Number_of_Borrowers)) +
  geom_bar(position = "dodge")
```

```{r}
# First Time Homebuyer & Conforming Flag
loan_data %>%
      filter(First_Time_Homebuyer_Flag %in% revised_hf) %>%
  ggplot(mapping = aes(x = First_Time_Homebuyer_Flag, fill = Super_Conforming_Flag)) +
  geom_bar(position = "dodge")
```
```{r}
# First Time Homebuyer & Risk Level
loan_data %>%
      filter(First_Time_Homebuyer_Flag %in% revised_hf) %>%
  ggplot(mapping = aes(x = First_Time_Homebuyer_Flag, fill = Risk_Level)) +
  geom_bar(position = "dodge")
```

```{r}
# Primary Occupancy Status & Property Type
occ_p <- c("P")

loan_data %>%
  filter(Occupancy_Status %in% occ_p) %>%
  ggplot(mapping = aes(x = Occupancy_Status, fill = Property_Type)) +
  geom_bar(position = "dodge")
```

```{r}
# Investment/Second Occupancy Status & Property Type
occ_is <- c("I", "S")
loan_data %>%
  filter(Occupancy_Status %in% occ_is) %>%
  ggplot(mapping = aes(x = Occupancy_Status, fill = Property_Type)) +
  geom_bar(position = "dodge")
```


```{r}
# Primary Occupancy Status & Loan Purpose
occ_p <- c("P")

loan_data %>%
  filter(Occupancy_Status %in% occ_p) %>%
  ggplot(mapping = aes(x = Occupancy_Status, fill = Property_Type)) +
  geom_bar(position = "dodge")
```

```{r}
# Investment/Second Occupancy Status & Loan Purpose
occ_is <- c("I", "S")
loan_data %>%
  filter(Occupancy_Status %in% occ_is) %>%
  ggplot(mapping = aes(x = Occupancy_Status, fill = Loan_Purpose)) +
  geom_bar(position = "dodge")
```

```{r}
# Primary Occupancy Status & Loan Purpose
occ_p <- c("P")

loan_data %>%
  filter(Occupancy_Status %in% occ_p) %>%
  ggplot(mapping = aes(x = Occupancy_Status, fill = Number_of_Borrowers)) +
  geom_bar(position = "dodge")
```

```{r}
# Investment/Second Occupancy Status & Loan Purpose
occ_is <- c("I", "S")
loan_data %>%
  filter(Occupancy_Status %in% occ_is) %>%
  ggplot(mapping = aes(x = Occupancy_Status, fill = Number_of_Borrowers)) +
  geom_bar(position = "dodge")
```

```{r}
# Primary Occupancy Status & Risk Level
occ_p <- c("P")

loan_data %>%
  filter(Occupancy_Status %in% occ_p) %>%
  ggplot(mapping = aes(x = Occupancy_Status, fill = Risk_Level)) +
  geom_bar(position = "dodge")
```

```{r}
# Investment/Second Occupancy Status & Risk Level
occ_is <- c("I", "S")
loan_data %>%
  filter(Occupancy_Status %in% occ_is) %>%
  ggplot(mapping = aes(x = Occupancy_Status, fill = Risk_Level)) +
  geom_bar(position = "dodge")
```


```{r}
# Loan Purpose & Number of Borrowers
loan_data %>%
  ggplot(mapping = aes(x = Loan_Purpose, fill = Number_of_Borrowers)) +
  geom_bar(position = "dodge")
```

```{r}
# Loan Purpose & Conforming Flag
loan_data %>%
  ggplot(mapping = aes(x = Loan_Purpose, fill = Super_Conforming_Flag)) +
  geom_bar(position = "dodge")
```


```{r}
# Number of borrowers & Risk Level
loan_data %>%
  ggplot(mapping = aes(x = Number_of_Borrowers, fill = Risk_Level)) +
  geom_bar(position = "dodge")
```

## Quantitative and Qualitative
+ Credit score.
+ CLTV.
+ DTI.

# Visuals
## Credit Score & Risk Level

```{r}
loan_data %>%
  ggplot(mapping = aes(x = Credit_Score, y = Risk_Level, fill = Risk_Level)) +
  geom_violin() +
  coord_flip() +
  theme_classic() +
  labs(title = "Low-risk borrowers sees high credit scores",
       subtitle = "High-risk group sees a narrower range in credit score") +
  ylab("Risk Level") +
  xlab("Credit Score") + 
  scale_fill_discrete(name = "Risk Level", labels = c("Low-risk", "High-risk")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right")
```

## Median DTI Ratio by Year & Risk Level 
```{r}
loan_data$Maturity_Date <- sub('(.{4})(.{2})', "\\1-\\2",loan_data$Maturity_Date)

```

```{r}
loan_data$Maturity_Date <- paste0(loan_data$Maturity_Date, "-01")

```

```{r}
# Transforming variables into dates
loan_data$Maturity_Date <- ymd(loan_data$Maturity_Date)
```

```{r}
complete.cases(loan_data$DTI_Ratio)
x <- loan_data[complete.cases(loan_data$DTI_Ratio), ]
x <- na.omit(loan_data)
```

```{r}
x %>%
  filter(State %in% "CA") %>%
  mutate(Year = year(Maturity_Date)) %>%
  group_by(Year, Risk_Level) %>%
  summarise(median_cltv = median(CLTV)) %>%
  ggplot(aes(x = Year, y = median_cltv, color = Risk_Level)) +
  geom_line() +
  theme_classic() +
  scale_color_discrete(name = "Risk Level", labels = c("Low-risk", "High-risk")) +
  labs(title = "Consistenly high CLTV ratio for high-risk borrowers in California",
       subtitle = "Low-risk borrowers see erratic patterns in CLTV ratio") +
  ylab("Median CLTV (in %)") +
  xlab("Maturity Year") + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right") 
```

```{r}
x %>%
  filter(State %in% "CA") %>%
  ggplot(aes(x =  DTI_Ratio, fill = Risk_Level)) +
  geom_boxplot() +
  coord_flip() +
  theme_classic() +
  scale_fill_discrete(name = "Risk Level", labels = c("Low-risk", "High-risk")) +
  labs(title = "Lower DTI ratio for low-risk borrowers in California",
       subtitle = "Low-risk borrowers see wide DTI ratio range and various anomalies") +
  xlab("Debt-to-Income Ratio (in %)") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "right") 
```

# Regression Tree

```{r}
library(rpart)
library(rpart.plot)
library(forecast)
library(caret)
```
```{r}
names(loan_data)
```

```{r}
loandata <- loan_data[ , c(2,4,6,8:28)]
names(loandata)
```

```{r}
head(loandata,10)
```

```{r}
set.seed(666)

train_index <- sample(1:nrow(loandata), 0.6 * nrow(loandata))
valid_index <- setdiff(1:nrow(loandata), train_index)

train_df <- loandata[train_index, ]
valid_df <- loandata[valid_index, ]

```

```{r}
str(loan_data)
```

```{r}
summary(loan_data)
```

```{r}
regress_tr <- rpart(Risk_Level ~ Credit_Score + MSA + Number_of_Units + DTI_Ratio + Original_UPB + Channel + PPM + Product_Type + State + Property_Type + Loan_Purpose + Original_Loan_Term, data = loandata, method = "anova")
prp(regress_tr)
```














